Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-14 paper.tex
Major refactor of introduction section

@@

% TODO: Reframe this paragraph to make it like a better background introduction.
As the OMOP CDM has become a cornerstone in running these widely distributed network studies, special attention to developing tools around handling OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.
It features a person-centric design where each domain records personal identity while prioritizing data protection through the limiting of information that could endanger patient anonymity.

% TODO: Reframe this to suggest the more general impact that OHDSI has had within the observational health research space.
Driving development within the subecosystem is the influence the OHDSI open-science network.
As mentioned previously, OHDSI is a global, collaborative, interdisciplinary, and open network of researchers, healthcare professionals, students, patients, and other entities committed to large-scale health data analytics and generating reliable evidence, with the goal of promoting better health decisions and better care. 
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
What could be considered the magnum opus of the OMOP group was the development of what is called the OMOP Common Data Model (OMOP CDM) to house and ingest this sort of data in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research.

@@

% TODO: Use this content to address what the notion of "network studies" are; move this to the "Introduction" section
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of the OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

@@

\footnotetext{I am unsure of where this notion came from. I believe it may have originally been coined by my former mentor and OHDSI founder, Dr. Jon Duke. 
The provenance of related terms like "OMOP-ification" (the process of converting a database to the OMOP CDM) is even murkier.}

\subsection{What Is JuliaHealth?}

@@

\subsection{Status of Conducting Observational Health Research in Julia}

% TODO: Move this to What Is JuliaHealth section
As JuliaHealth is an ad hoc network of contributors, varying aspects of the ecosystem have developed at differing rates as a function of factors such as developer time or workplace incentives.

>>>>

% TODO: Use this to address the notion of what "network studies" are
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of the OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

@@

\subsubsection{Core Technologies within Observational Health Research}

% TODO: Reframe this paragraph to make it like a better background introduction.
What could be considered the magnum opus of the OMOP group was the development of what is called the OMOP Common Data Model (OMOP CDM) to house and ingest this sort of data in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
As the OMOP CDM has become a cornerstone in running these widely distributed network studies, special attention to developing tools around handling OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.
It features a person-centric design where each domain records personal identity while prioritizing data protection through the limiting of information that could endanger patient anonymity.

\subsubsection{The Impact of the Observational Health Data Science and Informatics (OHDSI) Consortium}

% TODO: Reframe this to suggest the more general impact that OHDSI has had within the observational health research space.
Driving development within the subecosystem is the influence the OHDSI open-science network.
As mentioned previously, OHDSI is a global, collaborative, interdisciplinary, and open network of researchers, healthcare professionals, students, patients, and other entities committed to large-scale health data analytics and generating reliable evidence, with the goal of promoting better health decisions and better care. 
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research.

@@

\subsubsection{JuliaHealth's Role within Observational Health Research}

As JuliaHealth is an ad hoc network of contributors, varying aspects of the ecosystem have developed at differing rates as a function of factors such as developer time or workplace incentives.

@@

\footnotetext{It is unclear where terms like "OMOP-ify" or "OMOP-ification" (the process of converting a data to the OMOP CDM) originated; it may have been coined by Dr. Jon Duke.}

@@
